---
title: "Overview of BiocPkgTools"
author:
- name: Sean Davis
  affiliation: National Cancer Institute, National Institutes of Health, Bethesda, MD USA
  email: seandavi@gmail.com
- name: Vince Carey
  affiliation: Channing Lab, Brigham and Womens Hospital, Harvard University, Boston, MA USA
- name: Lori Shepherd
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY USA
- name: Martin Morgan
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY USA
package: BiocPkgTools
output:
  BiocStyle::html_document:
    toc: false
abstract: |
  Bioconductor has a rich ecosystem of metadata around packages, usage, and build status. This package is a simple collection of functions to access that metadata from R in a tidy data format. The goal is to expose metadata for data mining and value-added functionality such as package searching, text mining, and analytics on packages. 
vignette: |
  %\VignetteIndexEntry{Overview of BiocPkgTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

Bioconductor has a rich ecosystem of metadata around packages, usage, and build status. This package is a simple collection of functions to access that metadata from R in a tidy data format. The goal is to expose metadata for data mining and value-added functionality such as package searching, text mining, and analytics on packages. 

Functionality includes access to :

- Download statistics
- General package listing
- Build reports
- Package dependency graphs
- Vignettes

```{r init, include=FALSE}
library(knitr)
opts_chunk$set(warning = FALSE, message = FALSE, cache=FALSE)
```

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

# Build reports

The Bioconductor build reports are available online as HTML pages. 
However, they are not very computable.
The `biocBuildReport` function does some heroic parsing of the HTML
to produce a *tidy* data.frame for further processing in R. 

```{r}
library(BiocPkgTools)
head(biocBuildReport())
```

## Personal build report

Because developers may be interested in a quick view of their own
packages, there is a simple function, `problemPage`, to produce an HTML report of
the build status of packages matching a given author *regex*. The default is
to report only "problem" build statuses (ERROR, WARNING).

```{r}
problemPage()
```

When run in an interactive environment, the `problemPage` function 
will open a browser window for user interaction. Note that if you want
to include all your package results, not just the broken ones, simply 
specify `includeOK = TRUE`.

# Download statistics

Bioconductor supplies download stats for all packages. The `biocDownloadStats`
function grabs all available download stats for all packages in all
Experiment Data, Annotation Data, and Software packages. The results
are returned as a tidy data.frame for further analysis.

```{r}
head(biocDownloadStats())
```

The download statistics reported are for ***all available versions*** of a package.
There are no separate, publicly available statistics broken down by version.

# Package details

The R `DESCRIPTION` file contains a plethora of information regarding package
authors, dependencies, versions, etc. In a repository such as Bioconductor, these
details are available in bulk for all inclucded packages. The `biocPkgList` returns
a data.frame with a row for each package. Tons of information are avaiable, as 
evidenced by the column names of the results.

```{r}
bpi = biocPkgList()
colnames(bpi)
```

Some of the variables are parsed to produce `list` columns. 

```{r}
head(bpi)
```

As a simple example of how these columns can be used, extracting
the `importsMe` column to find the packages that import the 
`r Biocpkg("GEOquery")` package.


```{r}
require(dplyr)
bpi = biocPkgList()
bpi %>% 
    filter(Package=="GEOquery") %>%
    pull(importsMe) %>%
    unlist()
```

# Dependency graphs

The Bioconductor ecosystem is built around the concept of interoperability
and dependencies. These interdependencies are available as part of the
`biocPkgList()` output. The `BiocPkgTools` provides some convenience
functions to convert package dependencies to R graphs. A modular approach leads
to the following workflow.

1. Create a `data.frame` of dependencies using `buildPkgDependencyDataFrame`.
2. Create an `igraph` object from the dependency data frame using `buildPkgDependencyIgraph`
3. Use native `igraph` functionality to perform arbitrary network operations. 
Convenience functions, `inducedSubgraphByPkgs` and `subgraphByDegree` are available.
4. Visualize with packages such as `r CRANpkg("visNetwork")`.


## Working with dependency graphs

A dependency graph for all of Bioconductor is a starting place.

```{r}
library(BiocPkgTools)
dep_df = buildPkgDependencyDataFrame()
g = buildPkgDependencyIgraph(dep_df)
g
library(igraph)
head(V(g))
head(E(g))
```

See `inducedSubgraphByPkgs` and `subgraphByDegree` to produce
subgraphs based on a subset of packages.

See the igraph documentation for more detail on graph analytics, setting
vertex and edge attributes, and advanced subsetting.

## Graph visualization

The visNetwork package is a nice interactive visualization tool
that implements graph plotting in a browser. It can be integrated
into shiny applications. Interactive graphs can also be included in
Rmarkdown documents (see vignette)

```{r}
igraph_network = buildPkgDependencyIgraph(buildPkgDependencyDataFrame())
``` 

The full dependency graph is really not that informative to look at, though
doing so is possible. A common use case is to visualize the graph of
dependencies "centered" on a package of interest. In this case, I will 
focus on the `r Biocpkg("GEOquery")` package. 

```{r}
igraph_geoquery_network = subgraphByDegree(igraph_network, "GEOquery")
```

The `subgraphByDegree()` function returns all nodes and connections within
`degree` of the named package; the default `degree` is `1`.

The visNework package can plot `igraph` objects directly, but more flexibility
is offered by first converting the graph to visNetwork form.

```{r}
library(visNetwork)
data <- toVisNetworkData(igraph_geoquery_network)
```

The next few code chunks highlight just a few examples of the visNetwork
capabilities, starting with a basic plot.

```{r}
visNetwork(nodes = data$nodes, edges = data$edges, height = "500px")
```

For fun, we can watch the graph stabilize during drawing, best viewed
interactively.

```{r}
visNetwork(nodes = data$nodes, edges = data$edges, height = "500px") %>%
    visPhysics(stabilization=FALSE)
```

Add arrows and colors to better capture dependencies.

```{r}
data$edges$color='lightblue'
data$edges[data$edges$edgetype=='Imports','color']= 'red'
data$edges[data$edges$edgetype=='Depends','color']= 'green'

visNetwork(nodes = data$nodes, edges = data$edges, height = "500px") %>%
    visEdges(arrows='from') 
```

Add a legend.

```{r}
ledges <- data.frame(color = c("green", "lightblue", "red"),
  label = c("Depends", "Suggests", "Imports"), arrows =c("from", "from", "from"))
visNetwork(nodes = data$nodes, edges = data$edges, height = "500px") %>%
  visEdges(arrows='from') %>%
  visLegend(addEdges=ledges)
```

## Integration with `r Biocpkg("BiocViews")`

The `r Biocpkg("biocViews")` package is a small ontology of terms describing
Bioconductor packages. 

```{r biocViews}
library(biocViews)
data(biocViewsVocab)
biocViewsVocab
library(igraph)
g = igraph.from.graphNEL(biocViewsVocab)
library(visNetwork)
gv = toVisNetworkData(g)
visNetwork(gv$nodes, gv$edges, width="100%") %>%
    visIgraphLayout(layout = "layout_as_tree", circular=TRUE)
```


# Provenance

```{r}
sessionInfo()
```
